Studying measurement invariance in factor analysis models via stochastic processes

### 2011-02-04 MIpaper/par_instable.tex
Partial revision to simulation section
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@35 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

To employ the LR test for assessing measurement invariance, the data
are typically separated into a certain number of given subgroups where
the parameters are estimated separately (for ease of
  exposition, we assume no parameters are restricted to be equal
  across subgroups). Then, the sum of maximized likelihoods 
from the subgroups are compared with the original maximized full-sample likelihood
in a $\chi^2$~test. More specifically, for the special case of two subgroups,
the alternative $H_1^*$ from (\ref{eq:h1*}) with fixed and prespecified $\nu$ is adopted
and the null hypothesis $H_0$ from (\ref{eq:h0}) reduces to ${\bm \theta}^{(A)} = {\bm \theta}^{(B)}$.
The parameter estimated $\hat {\bm \theta}^{(A)}$ can then be obtained from the
observations $i = 1, \dots, m$, say, for which $v_i \le \nu$. Analogously,
$\hat {\bm \theta}^{(B)}$ is obtained by maximizing the likelihood for the
observations $i = m + 1, \dots, n$, for which $v_i > \nu$. The LR test statistic
for the given threshold $\nu$ is then
\begin{equation} \label{eq:lr}
  \mathit{LR}(\nu) ~=~ -2 \left[
         \ell(\hat {\bm \theta}; x_1, \dots, x_n)
   ~-~ \{\ell(\hat {\bm \theta}^{(A)}; x_1, \dots, x_m)
    +    \ell(\hat {\bm \theta}^{(B)}; x_{m+1}, \dots, x_n)\}
    \right],
\end{equation}
which has an asymptotic $\chi^2$ with degrees of freedom equal to the number
of parameters in~${\bm \theta}$.
\readme{You previously modified the above section so that $\theta$
  consists only of parameters that are ``doubled'' under the
  alternative.  This appears to not allow for situations where
  parameters are restricted to be equal across subgroups.  I am ok
  with this, but most applications of LR tests with these models have
  some parameters restricted to be equal across subgroups.  So I've
  added some detail in parentheses at the beginning of the above
  paragraph.}

@@

\section{Simulation}
% Simulations demonstrating power of test(s) at various
% sample sizes, types of invariance violations, ?
In this section, we conduct a simulation designed to
examine the tests' power and Type I error rates in the context
described previously (testing three factor loadings for measurement
invariance).  We 
examine the power and error rates of four tests: the double-max test, the
Cram\'{e}r-von Mises test, the max-LM test, and the traditional likelihood
ratio (difference) test involving nested 
multiple-group models.  The likelihood ratio test has an advantage over the
other three tests here, in that it requires advance definition of
$\eta$.  This test is useful for comparative purposes,
however, as it helps indicate the extent to which the other tests
``lose'' power due to $\eta$ being unknown.

\subsection{Method}
Simulated data were generated in a similar way to that of 
the Example section.
The data included verbal scales that violated measurement invariance
with respect to students' ages, with students 14--15 years of age
having smaller factor loadings than students 16--17 years of age.  
Simulation data were generated from the same general factor analysis
model (see Figure~\ref{fig:famod}), but sample size and magnitude
of measurement 
invariance violation was manipulated to examine power:
we examined power to detect 
invariance violations across three sample sizes ($n=100,
200, 500$) and three magnitudes of violations (factor loadings
for the younger students being 0, 1, 2, or 3 standard
errors\footnote{The standard errors used here were
  obtained by fitting
  the model to a sample of size 100, where the sample consisted of all
  ``older'' subjects.} below 
those for the older students).  The 0-standard error condition was
used to study Type I error rate.

For each sample size $\times$
magnitude combination, 500 datasets were generated and tested for
measurement invariance.  For the 0-standard error condition, we
generated 5,000 datasets because small proportions are more
 difficult to estimate precisely.  In each dataset, half the
students were 14--15 
years of age and half were 16--17 years of age.  As stated previously,
group
designations ($\eta$) were assumed to be
known for the likelihood ratio test.

\subsection{Results}
% MODIFY RESULTS TO DISCUSS ALL FOUR TESTS
% ALSO NEED 0-SE CONDITION
For each test statistic, power and Type I error rates for the twelve
sample size $\times$ magnitude combinations are displayed in
Table~\ref{tab:sim1}.  The 0-SE column lists Type I error rates; this
is the situation where the tests reject the hypothesis of measurement
invariance at $\alpha=.05$ when none exist.  This column 
shows that the Cram\'{e}r-von Mises test has accurate Type I error
rates at all sample sizes examined.
The double-max test, on the other hand, is conservative at
all sample sizes examined.  Finally, the likelihood ratio test is
slightly liberal at smaller sample sizes, with Type I error rates of
.07 and .06 at $n=100$ and $200$, respectively.  Type I error for the
likelihood ratio test at $n=500$ is accurate, however.

Focusing on power, the table includes many interesting results.
First, the likelihood ratio test has superior power over the stochastic
process-based tests at all sample sizes and invariance violations.
This is likely due to the fact that the likelihood ratio test had
extra information ($\eta$), whereas the other tests
were testing measurement invariance over all values
of $\eta$.  While power was lower for the tests developed in this
paper, it was not terribly low.  At $n=200$ and above, all tests were
on the same side of the traditional power cutoff of 0.8.  The tests
were able to detect large measurement invariance violations (of three
and sometimes two standard errors) at the larger sample sizes, and
they were generally unable to detect the one-standard-error violation
at any of the sample sizes studied.

Comparing the three proposed test statistics, we see that the
Cram\'{e}r-von Mises test and max-LM test are comparable, and they are
both better than the double-max test in all conditions.  This is
similar to the general recommendation described by Zeileis et al.\
\citeyear{ZeiSha10}, stating that tests involving sums of scores
are generally superior when multiple parameters change simultaneously.
Because the three verbal factor loadings all changed
across the two subgroups in the simulation, the Cram\'{e}r-von Mises
and max-LM tests excel.

In summary, we found that the proposed tests have less power than a
likelihood ratio test with known subgroups.  However, the power levels
remain adequate after considering the fact that the proposed tests are
testing measurement invariance across all possible subgroups.  For the
two-factor model examined here, all tests had trouble detecting small
measurement invariance violations at small sample size.  Larger
violations were more easily detected at larger sample sizes, with the
Cram\'{e}r-von Mises and max-LM tests having better power due to multiple
parameters being involved in the invariance violation.

% 0-SE column is fixed for n=100 (using 5000 reps instead of 500)
\begin{table*}
\caption{Simulated power and Type I error rates for four test statistics
  across three sample sizes and four magnitudes of measurement
  invariance violations. Abbreviations: CvM = Cram\'{e}r-von Mises
  test; MLM = Maximum Lagrange multiplier test; DM = Double-max test; 
  $\chi^2$ = Chi-square likelihood ratio test.}
\label{tab:sim1}
\begin{tabular}{llcccc} \thickline
 & & \multicolumn{4}{c}{Violation Magnitude (SEs)} \\
$n$ & Test Stat & 0 & 1 & 2 & 3 \\\hline
100 & CvM       & 0.05 & 0.10 & 0.33 & 0.63 \\ 
    & MLM       &      & 0.09 & 0.28 & 0.64 \\
    & DM        & 0.03 & 0.06 & 0.19 & 0.41 \\ 
    & $\chi^2$  & 0.07 & 0.16 & 0.49 & 0.83 \\ \hline
200 & CvM       & 0.05 & 0.19 & 0.64 & 0.96 \\
    & MLM       &      & 0.19 & 0.63 & 0.96 \\ 
    & DM        & 0.03 & 0.12 & 0.44 & 0.88 \\ 
    & $\chi^2$  & 0.06 & 0.27 & 0.76 & 0.99 \\ \hline
500 & CvM       & 0.05 & 0.44 & 0.98 & 1.00 \\ 
    & MLM       &      & 0.41 & 0.99 & 1.00 \\
    & DM        & 0.04 & 0.35 & 0.94 & 1.00 \\ 
    & $\chi^2$  & 0.05 & 0.58 & 1.00 & 1.00 \\\thickline
\end{tabular}
\end{table*}

>>>>

To employ the LR test for assessing measurement invariance, the data
are typically separated into a certain number of given subgroups where
the parameters are estimated separately (for ease of
  exposition, we assume no parameters are restricted to be equal
  across subgroups). Then, the sum of maximized likelihoods 
from the subgroups are compared with the original maximized full-sample likelihood
in a $\chi^2$~test. More specifically, for the special case of two subgroups,
the alternative $H_1^*$ from (\ref{eq:h1*}) with fixed and prespecified $\nu$ is adopted
and the null hypothesis $H_0$ from (\ref{eq:h0}) reduces to ${\bm \theta}^{(A)} = {\bm \theta}^{(B)}$.
The parameter estimated $\hat {\bm \theta}^{(A)}$ can then be obtained from the
observations $i = 1, \dots, m$, say, for which $v_i \le \nu$. Analogously,
$\hat {\bm \theta}^{(B)}$ is obtained by maximizing the likelihood for the
observations $i = m + 1, \dots, n$, for which $v_i > \nu$. The LR test statistic
for the given threshold $\nu$ is then
\begin{equation} \label{eq:lr}
  \mathit{LR}(\nu) ~=~ -2 \left[
         \ell(\hat {\bm \theta}; x_1, \dots, x_n)
   ~-~ \{\ell(\hat {\bm \theta}^{(A)}; x_1, \dots, x_m)
    +    \ell(\hat {\bm \theta}^{(B)}; x_{m+1}, \dots, x_n)\}
    \right],
\end{equation}
which has an asymptotic $\chi^2$ with degrees of freedom equal to the number
of parameters in~${\bm \theta}$.
\readme{Achim: You previously modified the above section so that $\theta$
  consists only of parameters that are ``doubled'' under the
  alternative.  This appears to not allow for situations where
  parameters are restricted to be equal across subgroups.  I am ok
  with this, but many applications of LR tests with these models have
  some parameters restricted to be equal across subgroups.  So I've
  added some detail in parentheses at the beginning of the above
  paragraph.}

@@

\section{Simulation}
% Simulations demonstrating power of test(s) at various
% sample sizes, types of invariance violations, ?
In this section, we conduct a simulation designed to
examine the tests' power and Type I error rates in a context
similar to the example.
We examine the power and error rates of three tests: the double-max test, the
Cram\'{e}r-von Mises test, and the max-LM test.  We also compare 
tests involving only the parameters that changed with tests of all model
parameters.  The example implied that the former tests had more
power, and the simulations provide more detail on the extent to which
power increases.  Finally, we also examine the tests' power across
various magnitudes of measurement invariance violations.

\subsection{Method}
Data were generated in a similar way to that of 
the Example section.
The data included verbal scales that violated measurement invariance
with respect to students' ages, with students 13--15 years of age
having smaller factor loadings than students 16--18 years of age.  
Simulation data were generated from the same general factor analysis
model (see Figure~\ref{fig:famod}), but sample size and magnitude
of measurement 
invariance violation were manipulated to examine power:
we examined power to detect 
invariance violations across three sample sizes ($n=100,
200, 500$) and three magnitudes of violations (factor loadings
for the younger students being $(0, 0.25, 0.5, \dots, 5)$ standard
errors\footnote{The standard errors used here were
  obtained by fitting
  the model to a sample of size 100, where the sample consisted of all
  ``older'' subjects.} below 
those for the older students).  The 0-standard error condition was
used to study Type I error rate.

\readme{Achim: Is it true that you used 5000 reps per cell?}
For each combination of sample size $\times$
violation magnitude $\times$ number of parameters being tested,
5,000 datasets were generated and tested.  In each dataset, half the
students were 13--15 
years of age and half were 16--18 years of age.

\subsection{Results}
% MODIFY RESULTS TO DISCUSS ALL FOUR TESTS
% ALSO NEED 0-SE CONDITION
Full simulation results are presented in Figure~\ref{fig:simres}, and
a subset of the results (for finer distinctions) is displayed in
Table~\ref{tab:simres}.  In describing the results, we largely refer
to the figure.

Figure~\ref{fig:simres} displays power as a function of violation
magnitude, with panels for each combination of sample size $\times$
number of parameters being tested.  
Circles reflect the double-max test, triangles reflect the max-LM
test, and plus signs reflect the Cram\`{e}r-von Mises test.  One can
generally observe that simultaneous tests of all 19 parameters
result in decreased power, with the tests performing very similarly at
the larger sample sizes.  The tests distinguish themselves from one
another when only the three factor loadings are tested, with the
Cram\`{e}r-von Mises test having the most power, followed by max-LM,
followed by the double-max test.  This advantage decreases with
increases in sample size.  
Absolute differences in power can be found in Table~\ref{tab:simres},
which contains a subset of the results.  For example, the table shows that the
power advantage of the Cram\`{e}r-von
Mises test can be as large as .1 (at $n=100$, 3 parameters being
tested).  It also shows that the Cram\`{e}r-von Mises test generally
has true Type-I error rates, with the double-max test being slightly
conservative and the max-LM test being slightly liberal at small
sample sizes.

In summary, we found that the proposed tests have adequate power to detect
measurement invariance violations in applied contexts.  The specific
statstics' are not equivalent to one another, and we found that the
Cram\`{e}r-von Mises statistic exhibited the best performance for the
data generated here.  In the discussion, we consider the extent to
which these findings may extend to other factor-analytic situations.

% Need to modify table and add figure
\begin{table*}
\caption{Simulated power and Type I error rates for four test statistics
  across three sample sizes and four magnitudes of measurement
  invariance violations. Abbreviations: CvM = Cram\'{e}r-von Mises
  test; MLM = Maximum Lagrange multiplier test; DM = Double-max test; 
  $\chi^2$ = Chi-square likelihood ratio test.}
\label{tab:sim1}
\begin{tabular}{llcccc} \thickline
 & & \multicolumn{4}{c}{Violation Magnitude (SEs)} \\
$n$ & Test Stat & 0 & 1 & 2 & 3 \\\hline
100 & CvM       & 0.05 & 0.10 & 0.33 & 0.63 \\ 
    & MLM       &      & 0.09 & 0.28 & 0.64 \\
    & DM        & 0.03 & 0.06 & 0.19 & 0.41 \\ 
    & $\chi^2$  & 0.07 & 0.16 & 0.49 & 0.83 \\ \hline
200 & CvM       & 0.05 & 0.19 & 0.64 & 0.96 \\
    & MLM       &      & 0.19 & 0.63 & 0.96 \\ 
    & DM        & 0.03 & 0.12 & 0.44 & 0.88 \\ 
    & $\chi^2$  & 0.06 & 0.27 & 0.76 & 0.99 \\ \hline
500 & CvM       & 0.05 & 0.44 & 0.98 & 1.00 \\ 
    & MLM       &      & 0.41 & 0.99 & 1.00 \\
    & DM        & 0.04 & 0.35 & 0.94 & 1.00 \\ 
    & $\chi^2$  & 0.05 & 0.58 & 1.00 & 1.00 \\\thickline
\end{tabular}
\end{table*}
